Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-03 paper/paper.tex
Finished edits for Section 3

\begin{figure}[h]
\begin{align*}
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & \Coloneqq && \textit{Terms} \\
            & \mid [[Set i]] \mid [[x]] \mid [[Void]] \mid [[absurd b]]
              && \textrm{universes, variables, empty type, explosion} \\
            & \mid [[Pi x : A . B]] \mid [[\ x . a]] \mid [[a b]]
              && \textrm{function types, abstractions, applications} \\
            & \mid [[a ~ b : A ]] \mid [[refl]] \mid [[J c p]]
              && \textrm{equality types, reflexivity proof, J eliminator} \\
            & \mid [[Bool]] \mid [[true]] \mid [[false]] \mid [[if a b0 b1]]
              && \textrm{boolean type, true, false, conditionals} \\
            % & \mid [[if a b0 b1]]
            %   && \textrm{conditional expression} \\
[[G]] & \Coloneqq [[empty]] \mid [[G ++ x : A]] && \textit{Typing contexts} \\
[[rho]] & \in [[SVar -> STm]] && \textit{Substitutions}
\end{align*}
\caption{Syntax of \lang}
\label{fig:syntax}
\end{figure}

@@

\subsection{Properties about the Logical Relation}
In the rest of this section, we develop the theory of our logical relation
with the goal of showing four key facts: irrelevance
(Lemma~\ref{lemma:logrelcoherence}), functionality
(Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure property
(Lemma~\ref{lemma:logrelbackclos}).  For the majority of the properties that
we prove in this section, we do not need any information about the
parameterized function $[[I]]$.  Each property about $[[InterpR i A S]]$
follows as a corollary of a property about $[[Interp I i A S]]$ with no or few
assumptions imposed on $[[I]]$. As a result, we usually state our lemmas in
terms of $[[Interp I i A S]]$ without duplicating them in terms of
$[[InterpR i A S]]$.

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi x
: A0 .B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl /\ a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
    I i Pi x : A . B S1]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi x : A . B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi x : A . B S1]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi x : A . B => Pi x : A0 . B0]]$ and $[[Interp I i Pi
    x : A0 . B0 S1]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a / x } F a #)]]$
    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi x : A . B => Pi x : A0 . B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ additionally satisfy the following properties.
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
    \end{itemize}
    These properties are exactly what we need to finish the proof.
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction too.
\begin{lemma}[Forward preservation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.

  The only interesting case is \rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B1]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Red} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

  The remaining cases all fall from induction hypotheses and basic
  properties about convertibility and parallel reduction we have
  established in Section~\ref{sec:spec}.
\end{proof}
From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
derive the following corollary that two convertible types can always interpret
into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
and refer to this property as irrelevance.
\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
  \label{lemma:logrelcoherence}
  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
\end{corollary}

Because the definition of our logical relation is an inductive relation,
it is not immediately obvious why each type $[[A]]$ can only uniquely
correspond to one set $[[S]]$. The following lemma shows that our
logical relation is indeed functional.
\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
  \label{lemma:logreldeter}
  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
\end{lemma}
\begin{proof}
  The proof proceeds by induction over the derivation of the first
  premise $[[Interp I i A S0]]$.
  All cases that are not \rref{I-Red} follow immediately from
  Lemma~\ref{lemma:interpinv}, the inversion properties.

  For \rref{I-Red}, we are given that there exists some $[[B]]$ such
  that $[[A => B]]$ and $[[Interp I i B S0]]$. Our goal is to show
  that given $[[Interp I i A S1]]$ for some $[[S1]]$, we have $[[S0 =
  S1]]$. By the preservation property
  (Lemma~\ref{lemma:interppreservation}),
  we know that $[[Interp I i B S1]]$ since $[[A => B]]$. The statement
  $[[S0 = S1]]$ then immediately follows from the induction hypothesis.
\end{proof}

Lemma~\ref{lemma:logreldeter} enables us to show the following
improved inversion lemma for function types whose statement is free of
the relation $[[F]]$, analogous to the derivable \rref{I-PiAlt}.
\begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a
    / x}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 implies forall
    S1, (# Interp I i B {a / x} S1 implies  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

The next lemma shows that our logical relation satisfies
cumulativity. That is, if a type has an interpretation at a lower
universe level, then we can obtain the same interpretation at a higher
universe level.
\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
  \label{lemma:logrelcumulativity}
  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
\end{lemma}
\begin{proof}
  Trivial by structural induction over the derivation of $[[Interp I
  i0 A S]]$.
\end{proof}
Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
implicitly assume that $[[I]]$ is defined on the set of natural
numbers less than $[[i1]]$.

\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
  \label{lemma:logreldeterhet}
  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
\end{corollary}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:logreldeter} and
  \ref{lemma:logrelcumulativity}.
\end{proof}

\begin{definition}[Sets closed under expansion]
  We say that a set of terms $[[S]]$ is closed under expansion if
  given $[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
\end{definition}
The final property we want to show is that the output set $[[S]]$ from
the logical relation is closed under expansion. Unlike the previous
lemmas, we directly state the lemma
in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
because we need to know something about $[[I]]$ for this property to
hold in the \rref{I-Set} case.

\yl{Can lift this lemma somewhere earlier in the section since it doesn't really
  depend on anything}
\scw{Jon didn't understand this proof}
\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
  \label{lemma:logrelbackclos}
  If $[[InterpR i A S]]$, then the set $[[S]]$ is closed under expansion.
\end{lemma}
\begin{proof}
  By the definition of $[[InterpR i A S]]$, we unfold $[[InterpR i A S]]$ by one
  step into $[[ Interp I i A S  ]]$ where $[[I j]] := [[{A | exists S
    , InterpR j A S}]]$.
  We then proceed by induction over the derivation of $[[Interp I i A S]]$.

  All cases are trivial except for the \rref{I-Set} case, where we
  want to show that the set $[[I j]]$ is closed under expansion for
  all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
  $[[A in I j]]$ if and only if there exists some $[[S]]$ such that
  $[[Interp I j A S]]$. By \rref{I-Red}, we must also have $[[B in I
  j]]$ for all $[[B => A]]$.
\end{proof}

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when
there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}

>>>>

\begin{figure}[h]
\begin{align*}
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]],[[C]] & \Coloneqq && \textit{Terms} \\
            & \mid [[Set i]] \mid [[x]] \mid [[Void]] \mid [[absurd b]]
              && \textrm{universes, variables, empty type, explosion} \\
            & \mid [[Pi x : A . B]] \mid [[\ x . a]] \mid [[a b]]
              && \textrm{function types, abstractions, applications} \\
            & \mid [[a ~ b : A ]] \mid [[refl]] \mid [[J c p]]
              && \textrm{equality types, reflexivity proof, J eliminator} \\
            & \mid [[Bool]] \mid [[true]] \mid [[false]] \mid [[if a b0 b1]]
              && \textrm{boolean type, true, false, conditionals} \\
            % & \mid [[if a b0 b1]]
            %   && \textrm{conditional expression} \\
[[G]] & \Coloneqq [[empty]] \mid [[G ++ x : A]] && \textit{Typing contexts} \\
[[rho]] & \in [[SVar -> STm]] && \textit{Substitutions}
\end{align*}
\caption{Syntax of \lang}
\label{fig:syntax}
\end{figure}

@@

\subsection{Properties of the Logical Relation}
In the rest of this section, we develop the theory of our logical relation
with the goal of showing four key facts: irrelevance
(Lemma~\ref{lemma:logrelcoherence}), functionality
(Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure property
(Lemma~\ref{lemma:logrelbackclos}).  For the majority of the properties that
we prove in this section, we do not need any information about the
parameterized function $[[I]]$.  Each property about $[[InterpR i A S]]$
follows as a corollary of a property about $[[Interp I i A S]]$ with no or few
assumptions imposed on $[[I]]$. As a result, we state most of our lemmas in
terms of $[[Interp I i A S]]$ without duplicating them in terms of
$[[InterpR i A S]]$.

First, we prove inversion principles for our logical relation. Given
$[[Interp I i A S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or
$[[Pi x : A . B]]$, the inversion lemma allows us to say something about the set
$[[S]]$. The proofs are simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.

\begin{lemma}[Inversion (logical relation)]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl /\ a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
    If $[[Interp I i Pi x : A . B S1]]$, then there exist
    $[[S in PowerSet STm]]$, $[[F in S -> PowerSet STm]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  We show only the inversion property for the function type.
  We start by induction on the derivation of $[[Interp I i Pi x : A . B S1]]$.
  There are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:]
    We are given that $[[Pi x : A . B => C]]$ and that $[[Interp I i C S1]]$.
    By inversion on parallel reduction, we have that $[[C]] = [[Pi x : A0 . B0]]$
    for some $[[A0]], [[B0]]$, and that $[[A => A0]]$ and $[[B => B0]]$.
    From the induction hypothesis on $[[Interp I i (# Pi x : A0 . B0 #) S1]]$,
    there exist $[[S in PowerSet STm]]$ and $[[F in S -> PowerSet STm]]$ such that:
    \begin{itemize}
      \item $[[Interp I i A0 S ]]$
      \item $[[forall a, (# a in S implies Interp I i B0 { a / x } F a #)]]$
      \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x}]]$ for all
    $[[a]]$. As a result, $[[S]]$ and $[[F]]$ satisfy the following properties
    by \rref{I-Red} on $[[A => A0]]$ and $[[B {a /x} => B0 {a/x}]]$, respectively:
    \begin{itemize}
      \item $[[Interp I i A S]]$
      \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
    \end{itemize}
    These properties are exactly what we need to finish the proof.
  \end{description}
  \vspace{-2\baselineskip}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction as well.

\begin{lemma}[Forward preservation (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}

\begin{proof}
  We carry out the proof by induction on the derivation of
  $[[Interp I i A S]]$.

  The only interesting case is \rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B1]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Red} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B1 S]]$.

  The remaining cases all fall from the induction hypotheses and the basic
  properties of convertibility and parallel reduction established in
  Section~\ref{sec:spec}.
\end{proof}

From forward preservation and \rref{I-Red}, we can easily
derive the following corollary that two convertible types always interpret
into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
and refer to this property as \emph{irrelevance}.

\begin{corollary}[Irrelevance (l.r.)\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
  \label{lemma:logrelcoherence}
  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
\end{corollary}

Because the definition of our logical relation is an inductive relation, it is
not immediately obvious that each type $[[A]]$ interprets to a unique set $[[S]]$.
The following lemma shows that our logical relation is indeed functional.

\begin{lemma}[Functionality (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
  \label{lemma:logreldeter}
  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
\end{lemma}

\begin{proof}
  The proof proceeds by induction on the derivation of the first
  premise $[[Interp I i A S0]]$.
  All cases that are not \rref{I-Red} follow immediately from
  Lemma~\ref{lemma:interpinv}, the inversion properties.

  In \rref{I-Red}, there exists some $[[B]]$ such that $[[A => B]]$ and
  $[[Interp I i B S0]]$. Our goal is to show that given $[[Interp I i A S1]]$
  for some $[[S1]]$, we have $[[S0 = S1]]$. By forward preservation
  (Lemma~\ref{lemma:interppreservation}) and $[[A => B]]$,
  we know that $[[Interp I i B S1]]$.
  Then $[[S0 = S1]]$ immediately follows from the induction hypothesis.
\end{proof}

Functionality allows us to prove an inversion lemma for the derivable
\rref{I-PiAlt}, which does not mention the function $[[F]]$ found in \rref{I-Pi}.

\begin{lemma}[Alternate inversion of function types (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}\leavevmode
  If $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
  such that:
  \begin{itemize}
    \item $[[Interp I i A S0]]$
    \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a / x} S1 #) #)]]$
    \item $[[S = { b | forall a, (# a in S0 implies forall S1, (# Interp I i B {a / x} S1 implies  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}

\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

The next lemma shows that the logical relation is cumulative:
if a type has an interpretation at a lower universe level,
then we can obtain the same interpretation at a higher universe level.

\begin{lemma}[Cumulativity (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
  \label{lemma:logrelcumulativity}
  If $[[Interp I i A S]]$ and $[[i < j]]$, then $[[Interp I j A S]]$.
\end{lemma}

\begin{proof}
  Trivial by structural induction over the derivation of $[[Interp I i A S]]$.
\end{proof}

Note that in the statement of cumulativity, we implicitly assume that $[[I]]$
is defined on naturals strictly less than $[[j]]$. Then by cumulativity and
trichotomy of the order on naturals (i.e. for any $i, j$, either $i < j$ or
$i = j$ or $i > j$), we show that functionality holds even when considering
interpretations of type at different universe levels.

\begin{corollary}[Level-irrelevant functionality (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
  \label{lemma:logreldeterhet}
  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
\end{corollary}

\begin{proof}
  Immediate from Lemmas~\ref{lemma:logreldeter} and
  \ref{lemma:logrelcumulativity}.
\end{proof}

The final property we want to show is that the output set $[[S]]$ from
the logical relation is closed under expansion. Unlike the previous
lemmas, we directly state the lemma
in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
because we need the actual instantiation of $I$ to prove the \rref{I-Set} case.

\begin{definition}[Closure under expansion]
  We say that a set of terms $[[S]]$ is closed under expansion if
  given $[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
\end{definition}

\begin{lemma}[Backward closure of interpretations (l.r.)\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
  \label{lemma:logrelbackclos}
  If $[[InterpR i A S]]$, then $[[S]]$ is closed under expansion.
\end{lemma}

\begin{proof}
  By the definition of $[[InterpR i A S]]$, we unfold $[[InterpR i A S]]$ into
  $[[Interp I i A S]]$ where $[[I j]] := [[{A | exists S, InterpR j A S}]]$ for
  $[[j < i]]$. We then proceed by induction on the derivation of
  $[[Interp I i A S]]$.

  All cases are trivial except for the \rref{I-Set} case, where the goal is to
  show that $[[I j]]$ is closed under expansion for all $[[j < i]]$;
  that is, if $[[B => A]]$ and $[[A in I j]]$, then $[[B in I j]]$.
  By the definition of $[[I]]$, this is equivalent to showing that if
  $[[B => A]]$ and $[[Interp I j A S]]$ for some $[[S]]$,
  then there exists some $[[S0]]$ such that $[[Interp I j B S0]]$.
  Letting $[[S0]]$ be $[[S]]$, by \rref{I-Red},
  we have that $[[Interp I j B S]]$ as required.
\end{proof}

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}

\scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when
there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}
